---
title: "Question3r"
output: html_document
---

Importing Libraries:

```{r}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(tidyr)
#install.packages("gridExtra")
#library(gridExtra)
```


reading data

```{r}

flight_06 <- read_csv("/Users/rak/Desktop/2006.csv")
flight_07 <- read_csv("/Users/rak/Desktop/2007.csv")
airports <- read_csv("/Users/rak/Desktop/airports.csv")
carriers <- read_csv("/Users/rak/Desktop/carriers.csv")
plane_data <- read_csv("/Users/rak/Desktop/plane-data.csv")
```

# Data Cleaning:
```{r}
cat(dim(flight_06), dim(flight_07))
```
```{r}
# Checking for null values
colSums(is.na(flight_06))
colSums(is.na(flight_07))
```
```{r}
# Rows with cancelled and diverted flights are dropped:
flight_06 <- flight_06 %>% filter(Cancelled == 0, Diverted == 0)
flight_07 <- flight_07 %>% filter(Cancelled == 0, Diverted == 0)
```
```{r}
# Checking null values again:
colSums(is.na(flight_06))
colSums(is.na(flight_07))
```

Merging dataset
```{r}
# Merging the Dataframes
flight_data <- bind_rows(flight_06, flight_07)
```

Dropping redundant columns
```{r}
# Removing redundant columns:
flight_data <- select(flight_data, -CancellationCode, -TaxiIn, -TaxiOut)
colSums(is.na(flight_data))
```

Calculating TotalDelay and isDelay:
```{r}
# Calculating the total delay - arrival + departure delay
flight_data$TotalDelay <- flight_data$ArrDelay + flight_data$DepDelay
flight_data$TotalDelay <- as.numeric(flight_data$TotalDelay)
# 'isDelay' to determine if flight is delayed or not
flight_data$isDelay <- ifelse(flight_data$TotalDelay > 0, 1, 0)
flight_data$isDelay <- as.numeric(flight_data$isDelay)
```

# Question 3:

Finding the top destinations:
```{r}
top_destinations <- flight_data %>%
  count(Dest) %>%
  arrange(desc(n)) %>%
  top_n(4) %>%
  select(Dest)

print(top_destinations)
```

```{r}
# Creating 6 unique pairs of Flight trips amongst the top Four destinations:
pairs <- list(c("ATL","ORD"), c('ATL','DFW'), c('ATL','LAX'), c('ORD','DFW'), c('LAX','ORD'), c('LAX','DFW'))

pair_dict <- list()
```

Creating a new dictionary to store the flight data for all the pairs:
```{r}
# Adding flight data of each of the pairs to the dictionary:
for (i in 1:6) {
  origin <- pairs[[i]][1]    # Selecting origin
  destination <- pairs[[i]][2]    # Selecting destination

  # Combining both origin to destination and destination to origin as one pair
  PAIR1 <- flight_data %>%
    filter(Dest == origin, Origin == destination)
  PAIR2 <- flight_data %>%
    filter(Dest == destination, Origin == origin)
  PAIR <- bind_rows(PAIR1, PAIR2)
  pair_dict[[i]] <- PAIR
}
```

```{r}
# Changing the key number to the names of the pairs:
pair_dict[['ATL-ORD']] <- pair_dict[[1]]
pair_dict[['ATL-DFW']] <- pair_dict[[2]]
pair_dict[['ATL-LAX']] <- pair_dict[[3]]
pair_dict[['ORD-DFW']] <- pair_dict[[4]]
pair_dict[['LAX-ORD']] <- pair_dict[[5]]
pair_dict[['LAX-DFW']] <- pair_dict[[6]]


for (key in names(pair_dict)) {
  if (!is.null(pair_dict[[key]])) {
    pair_dict[[key]] <- pair_dict[[key]] %>%
      select(Year, Origin, Dest, Month)
  }
}
```

Grouping the data by month to find monthly data for each pair:
```{r}
#Creating new list to store the number of flight in each month for all the pairs
month_group <- list(
ATL_ORD = NULL,
ATL_DFW = NULL,
ATL_LAX = NULL,
ORD_DFW = NULL,
LAX_ORD = NULL,
LAX_DFW = NULL
)
```

```{r}
#Grouping by 'Month' to find the number of flights:
for (key in names(pair_dict)) {
month_group[[key]] <- table(pair_dict[[key]]$Month)
} 
```

Changing the dictionary of monthly flights to individual dataframes to allow the data to be graphed:

```{r}
df_ATL_ORD <- month_group[['ATL-ORD']]
df_ATL_DFW <- month_group[['ATL-DFW']]
df_ATL_LAX <- month_group[['ATL-LAX']]
df_ORD_DFW <- month_group[['ORD-DFW']]
df_LAX_ORD <- month_group[['LAX-ORD']]
df_LAX_DFW <- month_group[['LAX-DFW']]

# Convert tables to data frames
df_ATL_ORD <- as.data.frame(month_group[['ATL-ORD']])
df_ATL_DFW <- as.data.frame(month_group[['ATL-DFW']])
df_ATL_LAX <- as.data.frame(month_group[['ATL-LAX']])
df_ORD_DFW <- as.data.frame(month_group[['ORD-DFW']])
df_LAX_ORD <- as.data.frame(month_group[['LAX-ORD']])
df_LAX_DFW <- as.data.frame(month_group[['LAX-DFW']])

# Changing the column names to months and fligths:
colnames(df_ATL_ORD) <- c("Month", "Flights")
colnames(df_ATL_DFW) <- c("Month", "Flights")
colnames(df_ATL_LAX) <- c("Month", "Flights")
colnames(df_ORD_DFW) <- c("Month", "Flights")
colnames(df_LAX_ORD) <- c("Month", "Flights")
colnames(df_LAX_DFW) <- c("Month", "Flights")

df_ATL_ORD 
df_ATL_DFW 
df_ATL_LAX 
df_ORD_DFW 
df_LAX_ORD 
df_LAX_DFW 
```

Grouping the data by year to find yearly data for each pair:
```{r}
#Creating new list to store the number of flight in each year for all the pairs
year_group <- list(
ATL_ORD = NULL,
ATL_DFW = NULL,
ATL_LAX = NULL,
ORD_DFW = NULL,
LAX_ORD = NULL,
LAX_DFW = NULL
)
for (key in names(pair_dict)) {
year_group[[key]] <- table(pair_dict[[key]]$Year)
}
```

Changing the dictionary of yearly flights to individual dataframes to allow the data to be graphed:
```{r}
df_ATL_ORD_y <- year_group[['ATL-ORD']]
df_ATL_DFW_y <- year_group[['ATL-DFW']]
df_ATL_LAX_y <- year_group[['ATL-LAX']]
df_ORD_DFW_y <- year_group[['ORD-DFW']]
df_LAX_ORD_y <- year_group[['LAX-ORD']]
df_LAX_DFW_y <- year_group[['LAX-DFW']]

# Convert tables to data frames
df_ATL_ORD_y <- as.data.frame(year_group[['ATL-ORD']])
df_ATL_DFW_y <- as.data.frame(year_group[['ATL-DFW']])
df_ATL_LAX_y <- as.data.frame(year_group[['ATL-LAX']])
df_ORD_DFW_y <- as.data.frame(year_group[['ORD-DFW']])
df_LAX_ORD_y <- as.data.frame(year_group[['LAX-ORD']])
df_LAX_DFW_y <- as.data.frame(year_group[['LAX-DFW']])

# Changing the column names to months and fligths:
colnames(df_ATL_ORD_y) <- c("Year", "Flights")
colnames(df_ATL_DFW_y) <- c("Year", "Flights")
colnames(df_ATL_LAX_y) <- c("Year", "Flights")
colnames(df_ORD_DFW_y) <- c("Year", "Flights")
colnames(df_LAX_ORD_y) <- c("Year", "Flights")
colnames(df_LAX_DFW_y) <- c("Year", "Flights")

df_ATL_ORD_y 
df_ATL_DFW_y
df_ATL_LAX_y 
df_ORD_DFW_y 
df_LAX_ORD_y 
df_LAX_DFW_y 
```

# Plotting monthly flights for each individual pair:
```{r}
plot(df_ATL_ORD$Flights, type = 'l', xaxt = 'n', xlab = "Months", ylab = "Monthly Flights", main="Monthly flights between ATL and ORD") 
axis(1, at = 1:12)  
plot(df_ATL_DFW$Flights, type = 'l', xaxt = 'n', xlab = "Months", ylab = "Monthly Flights", main="Monthly flights between ATL and DFW")  
axis(1, at = 1:12)  
plot(df_ATL_LAX$Flights, type = 'l', xaxt = 'n', xlab = "Months", ylab = "Monthly Flights", main="Monthly flights between ATL and LAX")  
axis(1, at = 1:12)  
plot(df_ORD_DFW$Flights, type = 'l', xaxt = 'n', xlab = "Months", ylab = "Monthly Flights", main="Monthly flights between ORD and DFW")  
axis(1, at = 1:12)  
plot(df_LAX_ORD$Flights, type = 'l', xaxt = 'n', xlab = "Months", ylab = "Monthly Flights", main="Monthly flights between LAX and ORD")  
axis(1, at = 1:12)  
plot(df_LAX_DFW$Flights, type = 'l', xaxt = 'n', xlab = "Months", ylab = "Monthly Flights", main="Monthly flights between LAX and DFW") 
axis(1, at = 1:12) 
```

# Plotting monthly flights of all pairs:
```{r}
# Create empty plot
plot(0, type='n', xlim=c(1,12), ylim=c(0, 4000),
     xlab='Months', ylab='Monthly Flights',
     main='Monthly Flights for all pairs:')

# Add lines for each pair
lines(df_ATL_ORD$Flights, col='red')
lines(df_ATL_DFW$Flights, col='green')
lines(df_ATL_LAX$Flights, col='blue')
lines(df_ORD_DFW$Flights, col='orange')
lines(df_LAX_ORD$Flights, col='purple')
lines(df_LAX_DFW$Flights, col='brown')

# Add legend
legend('topright', legend=c('ATL-ORD', 'ATL-DFW', 'ATL-LAX', 'ORD-DFW', 'LAX-ORD', 'LAX-DFW'), 
       col=c('red', 'green', 'blue', 'orange', 'purple', 'brown'), lty=1, cex=0.8)
```

# Plotting percentage change in monthly flights for all pairs:

```{r}
# Calculating percentage change in month;y flights for each pair:
df_ATL_ORD_pc <- diff(df_ATL_ORD$Flights)/lag(df_ATL_ORD$Flights) * 100
df_ATL_DFW_pc <- diff(df_ATL_DFW$Flights)/lag(df_ATL_DFW$Flights) * 100
df_ATL_LAX_pc <- diff(df_ATL_LAX$Flights)/lag(df_ATL_LAX$Flights) * 100
df_ORD_DFW_pc <- diff(df_ORD_DFW$Flights)/lag(df_ORD_DFW$Flights) * 100
df_LAX_ORD_pc <- diff(df_LAX_ORD$Flights)/lag(df_LAX_ORD$Flights) * 100
df_LAX_DFW_pc <- diff(df_LAX_DFW$Flights)/lag(df_LAX_DFW$Flights) * 100
```

```{r}
plot(0, type = 'n', xlim = c(2, 12), ylim = c(-50, 50),
     xlab = 'Months', ylab = 'Percentage Change in Monthly Flights',
     main = 'Percentage Change in Monthly Flights for all pairs')

# Add lines for each pair
lines(df_ATL_ORD_pc, col = 'red')
lines(df_ATL_DFW_pc, col = 'green')
lines(df_ATL_LAX_pc, col = 'blue')
lines(df_ORD_DFW_pc, col = 'orange')
lines(df_LAX_ORD_pc, col = 'purple')
lines(df_LAX_DFW_pc, col = 'brown')

# Add x-axis labels
axis(1, at = 1:12, labels = month.abb)

# Add legend
legend('topright', legend = c('ATL-ORD', 'ATL-DFW', 'ATL-LAX', 'ORD-DFW', 'LAX-ORD', 'LAX-DFW'), 
       col = c('red', 'green', 'blue', 'orange', 'purple', 'brown'), lty = 1, cex = 0.8)
```

# Plotting yearly flights for each of the pairs:
```{r}
df_year <- data.frame(
  Pair = rep(c("ATL-ORD", "ATL-DFW", "ATL-LAX", "ORD-DFW", "LAX-ORD", "LAX-DFW"), each = 2),
  Year = rep(c(2006, 2007), times = 6),
  Flights = c(df_ATL_ORD_y$Flights, df_ATL_DFW_y$Flights, df_ATL_LAX_y$Flights, 
              df_ORD_DFW_y$Flights, df_LAX_ORD_y$Flights, df_LAX_DFW_y$Flights)
)

# Creating a grouped bar chart
ggplot(df_year, aes(x = Pair, y = Flights, fill = factor(Year))) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  scale_fill_manual(values = c("#ff7f0e", "#2ca02c")) +
  labs(title = "Number of flights across different years", x = "Pairs", y = "Number of flights") +
  theme_minimal()
```








